---
title: "finalproject"
output: html_document
date: "2023-11-30"
---
```{r}
library(tidyverse)
```
# Final Project (Population vs. Obesity by State)

## Intro
### Orignial Questions: 
### 1. Are there any correlations between a state’s population size and its obesity percentage? For example, do more populous states tend to have higher or lower obesity rates?


### 2. What is the state with the highest obesity percentage, and how does its population compare to other states with lower obesity rates? Is there a relationship between high obesity and a larger or smaller population?


### 3. Are there any significant changes in state obesity percentages over time, and how do these changes relate to changes in state populations? 


### 4.Are there any states where obesity has been increasing or decreasing at a faster rate compared to changes in population size?


## Data cleaning
### Read in our data and clean population data for 2010-2019
```{r}
pop_2010_2019 <- read_delim("../Data/pop_2010_2019.csv")
pop_2010_2019

colnames(pop_2010_2019)[1] <- "geographic_area"
colnames(pop_2010_2019)[2] <- "census"
colnames(pop_2010_2019)[3] <- "est_base"
colnames(pop_2010_2019)[4] <- "pop_est"
colnames(pop_2010_2019)[9] <- 2015
colnames(pop_2010_2019)[10] <- 2016
colnames(pop_2010_2019)[11] <- 2017
colnames(pop_2010_2019)[12] <- 2018
colnames(pop_2010_2019)[13] <- 2019

short_pop_10_19 <- subset(pop_2010_2019, select = c("geographic_area", "2015", "2016", "2017", "2018", "2019"))
short_pop_10_19 <- short_pop_10_19[-c(1:3), ]
short_pop_10_19 <- short_pop_10_19[-c(57:58), ]
colnames(short_pop_10_19)[1] <- "area"

```

### Read in our data and clean population data for 2020-2022
```{r}
pop_2020_2022 <- read_delim("../Data/pop_2020_2022.csv")
pop_2020_2022

short_pop_20_22 <- subset(pop_2020_2022, select = c("NAME", "POPESTIMATE2020", "POPESTIMATE2021", "POPESTIMATE2022"))
short_pop_20_22
colnames(short_pop_20_22)[1] <- "area"
colnames(short_pop_20_22)[2] <- 2020
colnames(short_pop_20_22)[3] <- 2021
colnames(short_pop_20_22)[4] <- 2022

```

### Merging population data
```{r}
mergedpop <- left_join(short_pop_10_19, short_pop_20_22)
mergedInfo
```

### Read in our data and clean obesity data for 2015-2022
```{r}
obesity_data <- read_delim("../Data/National_Obesity_By_State.csv")
obesity_data
short_obesity_data <- subset(obesity_data, select = c("NAME", "Obesity", "Year"))
short_obesity_data

colnames(short_obesity_data)[1] <- "area"

longObesity <- short_obesity_data%>%
  pivot_wider(names_from = Year, values_from = Obesity)

longObesity <- longObesity[-c(39), ]

colnames(longObesity)[2] <- "ob_2015"
colnames(longObesity)[3] <- "ob_2016"
colnames(longObesity)[4] <- "ob_2017"
colnames(longObesity)[5] <- "ob_2018"
colnames(longObesity)[6] <- "ob_2019"
colnames(longObesity)[7] <- "ob_2020"
colnames(longObesity)[8] <- "ob_2021"
colnames(longObesity)[9] <- "ob_2022"


```

### Merging population data and obesity data by state
```{r}
mergedObesityPop <- left_join(longObesity, mergedpop)
mergedObesityPop
```

This is our dataset, population data from 2015-2022 and obesity percentages per state: **mergedObesityPop**. 

## Questions & Data
### 1. Are there any correlations between a state’s population size and its obesity percentage? For example, do more populous states tend to have higher or lower obesity rates?
```{r}

```


### 2. What is the state with the highest obesity percentage, and how does its population compare to other states with lower obesity rates? Is there a relationship between high obesity and a larger or smaller population?
```{r}
max_ob_2015 <- max(mergedObesityPop$ob_2015, na.rm = TRUE)
max_ob_2016 <- max(mergedObesityPop$ob_2016, na.rm = TRUE)
max_ob_2017 <- max(mergedObesityPop$ob_2017, na.rm = TRUE)
max_ob_2018 <- max(mergedObesityPop$ob_2018, na.rm = TRUE)
max_ob_2019 <- max(mergedObesityPop$ob_2019, na.rm = TRUE)
max_ob_2020 <- max(mergedObesityPop$ob_2020, na.rm = TRUE)
max_ob_2021 <- max(mergedObesityPop$ob_2021, na.rm = TRUE)
max_ob_2022 <- max(mergedObesityPop$ob_2022, na.rm = TRUE)
max_ob_2015
max_ob_2016 
max_ob_2017 
max_ob_2018 
max_ob_2019 
max_ob_2020
max_ob_2021
max_ob_2022 

maxOb_state_2015 <- mergedObesityPop$area[mergedObesityPop$ob_2015 == max_ob_2015]
maxOb_state_2016 <- mergedObesityPop$area[mergedObesityPop$ob_2016 == max_ob_2016]
maxOb_state_2017 <- mergedObesityPop$area[mergedObesityPop$ob_2017 == max_ob_2017]
maxOb_state_2018 <- mergedObesityPop$area[mergedObesityPop$ob_2018 == max_ob_2018]
maxOb_state_2019 <- mergedObesityPop$area[mergedObesityPop$ob_2019 == max_ob_2019]
maxOb_state_2020 <- mergedObesityPop$area[mergedObesityPop$ob_2020 == max_ob_2020]
maxOb_state_2021 <- mergedObesityPop$area[mergedObesityPop$ob_2021 == max_ob_2021]
maxOb_state_2022 <- mergedObesityPop$area[mergedObesityPop$ob_2022 == max_ob_2022]
maxOb_state_2015 
maxOb_state_2016 
maxOb_state_2017 
maxOb_state_2018
maxOb_state_2019
maxOb_state_2020
maxOb_state_2021
maxOb_state_2022 

```

Highest obesity percentage per year: 
2015: Louisiana
2016: West Virginia
2017: West Virginia
2018: West Virginia & Mississippi (tied)
2019: Mississippi
2020: Mississippi
2021: West Virginia
2022: West Virginia


### 3. Are there any significant changes in state obesity percentages over time, and how do these changes relate to changes in state populations? 
```{r}

```


### 4.Are there any states where obesity has been increasing or decreasing at a faster rate compared to changes in population size?
```{r}

```

