---
title: "finalproject"
output: html_document
date: "2023-11-30"
---
```{r, echo = FALSE}
library(tidyverse)
```
# Final Project (Population vs. Obesity by State)

**1. Introduction**		
1a) We are looking at the percentage of a state that is considered obese, to evaluate the correlation of obesity and population size. We know that America is viewed as more obese compared to foreign counterparts, and we know that some states have higher populations. We are interested to see if states with higher populations, or political positions affect obesity numbers. We want to view the connection between obesity numbers and population numbers to put it into perspective.
We are creating a dataset that combines the obesity percentage per state and the population data in that state to find patterns and trends between the size of the population and how obese they are in the USA. We also will use this data to look at the regions that contain a more obese population. Does the location of the state correlate with the percentage of the population that is obese?

1b) From the World Health Organization, we know that obesity numbers are only growing, practically tripling! This organization points out that no longer are upper income areas the only ones struggling with obesity. Lower and middle class areas are also struggling to keep obesity numbers low. More people are dying from being overweight than malnutrition which is completely preventable. World health even mentions that urban areas are struggling the most with obesity, which is something we hope to learn more about based on region and population size with our data. There are numerous diseases that are more prevalent due to obesity. If we can take a look at these numbers, focus on area hotspots, and raise awareness then we have a chance to keep population numbers high and obese deaths low. 
	We know that America is viewed as more obese compared to foreign counterparts, and we know that some states have higher populations. We are interested to see if states with higher populations affect obesity numbers. We want to add the connection between obesity numbers and population numbers to put it all into perspective.

1c) We would like to focus in on 4 main questions to be answered: 
What is the highest and lowest obesity state in 2015? How did that progress to 2022?
Are there any correlations between a state’s population size and its obesity percentage? For example, do more populous states tend to have higher or lower obesity rates?
Are there any significant changes in state obesity percentages over time, and how do these changes relate to changes in state populations? 
Are there any states where obesity has been increasing or decreasing at a faster rate? compare it to changes in population size?

1d) Obesity can lead to other health problems such as chronic diseases. Knowing what areas have higher obesity rates can help with narrowing down target intervention for health professionals or public workers. Answering our questions is interesting because knowing how population sizes relate to obesity offers insights into societal impact. It can help healthcare systems and later on increase productivity. Also, identifying patterns across states assists in customizing interventions for specific regions, effectively tackling disparities. Learning the changes of obesity alongside population dynamics provides foresight into forthcoming health concerns, steering proactive strategies. Overall, the results we find should serve as a tool for policymakers, aiding resource distribution and shaping effective public health strategies.


**2. Explain your Data**	
2a) We use these datasets due to their comprehensive nature and credibility. It has all the data we want as it includes obesity rates for each state over the years and also has the population divided for each state over the years. 
These datasets also come from credible sources, who collected the data for the purpose of showing the percentage of obesity per state, as well as the population of a state for future purposes. 
Basic summary statistics of final dataset (finalData): 104 rows, 11 columns, 8 total NA values for Puerto Rico (population value 2016-2022, growth_poc). 

2b) We found our obesity data from Data.gov, as well as the State of Childhood Obesity site, which pulls data from the same census as the data from Data.gov (https://stateofchildhoodobesity.org/demographic-data/adult/). Both obesity data originate from the data collected by the Behavioral Risk Factor Surveillance System (BRFSS) data, from the CDC. 
We found our population data from the United States Census Bureau website. The two data sets are from  (https://www.census.gov/data/datasets/time-series/demo/popest/2010s-state-total.html) and (https://www.census.gov/data/tables/time-series/demo/popest/2020s-state-total.html).

2c) For obesity data, the State of Childhood Obesity, sponsored by the Robert Wood Johnson Foundation, collected the data from the Behavioral Risk Factor Surveillance System (BRFSS) data, collected by the Centers for Disease Control and Prevention (aka. CDC).  
Both population data was gathered and compiled by the US Census Bureau. US Census Bureau collects data through a variety of surveys, such as the decennial census (conducted every 10 years) and other regular surveys like the American Community Survey (ACS), Economic Census, and others.  

2d) All of our data is publicly available and anyone can access it. The State of Childhood Obesity site has clearly organized the BRFSS data into a readable, easy to understand format. The US Census Bureau also has publicly posted the data into multiple formats so that it could be easily used by data analysis. 

2e) The population data for the years 2010 - 2019 is about state population totals from 2010 - 2019. It includes the population, population change, estimated components of population change from April, 2010 - July 1, 2019. The population data for years 2020 - 2022 contains similar information. The dataset is about state population totals and component change in years 2020 - 2022. Both data sets tell us the population for each State of the United States. 
The adult obesity data shows the percentage of the state by population that are considered obese from the years 2015 (and some earlier), up to 2022. The site which the dataset came from also has other measurable factors such as demographic, adult and child, as well as comprehensive overview and fun facts pertaining to obesity per state. Some include different age ranges. We use the adult obesity dataset.

2f) For the obesity data, we encountered some missing data, for example the obesity data for Puerto Rico, as for some data collection, Puerto Rico is not considered a state of the U.S., but a territory, and treated as different in some datasets.  
For the population data, we could not find one where it had years from 2015 - 2022 all in one data. Therefore, we had to use two separate population data, one that includes population of years 2010 - 2019 and one for the years 2020 - 2022. 
Our final dataset merge these two datasets, population and obesity together. We first merged the two population data sets (2010-2019 and 2020-2022) and cleaned. Then we merged the obesity and population data using the connector “Area” which is the state name. Only data quality issue was the missing obesity values for Puerto Rico. 

2g) In our final dataset, _finalData_,  each row corresponds to a U.S. state, providing information on its population size or obesity rate, with data categorized by type, “Obesity” or “Population” for the years spanning from 2015 to 2022. Additionally with an additional column of growth/rate of change (percentage change) of obesity and population for years between 2015-2022. Our final summarizing data frame is called _summaryDf_. 

2h) Following the processes of data cleaning and merging, we now have a cumulative total of 1,144 observations. Within the final dataset, there are a total of 104 rows, and 11 columns. 

2i) The relevant variables are Area, type, years 2015-2022, and the growth/rate of change columns. They are coded as their own columns, with values in each year and are classified by the _type_ column, which indicates if the value is correlated to obesity or population data. 

2j) There are a total of 8 missing values ("NA"): Puerto Rico has missing obesity percentage values for years 2016-2022 as well as the growth_pc. 


**3. Discuss your methods**		
3a)
Question 1: 
In order to tackle a trend heavy question, we started by searching the data for the states with the highest and lowest obesity percentages. Since the data we are working from reflects self reported obesity percentages, we knew that the states were aware of their obesity percentage. We focused on implementing data including the variables “Area”, “Year”, and “Obesity”. We were interested in finding how, in knowing whether they had a high or low obesity percentage, states were able to increase or decrease their percentage. 

Question 2: 
In addressing this specific question, we have employed all the variables at our disposal in the dataset, namely Area, Year, Population, and Obesity. The comprehensive utilization of all variables is imperative for our analysis, as our objective is to assess the states with the highest and lowest obesity rates each year. Subsequently, we aim to leverage these findings to examine the population dynamics of each state during those specific years.

Question 3: 
In answering this question, we first analyzed the changes in obesity rates and population. To do this we calculated changes in obesity rates for each state (obesity_changes) by subtracting the first recorded obesity rate from the last recorded rate. Then, we determined changes in population for each state (population_changes) by subtracting the first recorded population from the last recorded population. Using this we then identify any outliers in obesity rate changes by detecting outliers in obesity rate changes using a threshold of absolute scaled difference greater than 2 standard deviations (obesity_outliers). Finally, to compare between population changes and obesity outliers we used an inner join (inner_join) between population changes and obesity outliers to identify states with significant changes in obesity rates compared to their population changes. 

Question 4: 
For this specific question, I utilized multiple variables and forms of our data, transforming it into wide and long forms, and selecting variables targeted to specific aspects of the question. I employed variables such as Year (specifically 2022 and 2015), Obesity, Population, and Area. I additionally created two-three additional columns, “type” and “ob_growth” and “pop_growth”, these two which can be found in the “growth_roc” column in our final df “finalData.” I found the states which have the highest and lowest rate of change in obesity and population separately using these variables and arranging them (desc()), to see if there was a correlation between high obesity rates of change and high population rates of change. 
I utilized the tools such as ggolot to plot the relationship and visualize the highest/lowest rates of change for obesity and population. 

3b)
Question 1: 
In order to tackle the first question, we started by processing obesity data for the year 2015 and analyzing trends in obesity percentages for the years 2015 to 2022. We used dollar notation to access information and extract data for the year 2015 from the “finalMerge” dataset. We then used that data to find the states with the highest and lowest obesity percentage in 2015 utilizing the min() and max() feature. In order to find the annual trends from 2015 to 2022, we defined a function to calculate annual changes in obesity percentages, and then applied it to data for the states with the highest and lowest obesity. In order to provide accurate visualization to discuss the results, we created a graph, plotting the Obesity percentages from the states of Louisiana and Colorado and provided a trend line of their average increase/decrease throughout the years. This was to show us how the states responded to knowing their obesity percentages, and reflect whether there was improvement. 

Question 2: 
We initiate the process by identifying states with the highest obesity rates each year. Using the "finalmerge" dataset grouped by the "Year" variable, we employ the slice_max function to extract the row with the maximum value in the "Obesity" variable for each year. This step pinpoints the state with the highest obesity rate annually. After removing the grouping with the ungroup function, we store this information in a new dataframe named "high." Replicating this process, we also identify states with the lowest obesity rates each year, saved as the "low" dataframe. To facilitate a comprehensive comparison between the highest and lowest obesity rates and their correlation with population size, we merge these datasets using full_join based on common variables (Area, Year, Population, and Obesity), resulting in the "merge_high_low" dataset. Subsequently, through ggplot code blocks, we create visualizations, including a scatter plot and line chart, to depict obesity rates and population growth over the years. These visualizations, differentiating states by color, offer insights into the dynamics between high and low obesity rates and population changes. The comparison of these graphs allows for a deeper exploration of potential correlations between obesity rates and population sizes.

Question 3:
For this question we start the process by calculating obesity rate changes. We calculated changes in obesity rates for each state (variable: obesity_changes): Grouped the merged dataset (finalMerge) by "Area" (state) to compute the difference between the last recorded obesity rate and the first recorded obesity rate for each state. Then we calculated population changes: Calculated changes in population for each state (variable: population_changes): Grouped the merged dataset (finalMerge) by "Area" to calculate the difference between the last recorded population and the first recorded population for each state. Next, we identified obesity outliers:  Detected potential outliers in obesity rate changes (variable: obesity_outliers): Identified states where the absolute scaled change in obesity rates was greater than 2 standard deviations from the mean change. Finally, we compared population changes with obesity outliers: Merged the population changes with the identified obesity outliers (variable: population_vs_obesity_outliers): Performed an inner join between the population changes and obesity outliers by the common key "Area" to explore potential associations between significant obesity rate changes and corresponding population changes.

Question 4: 
We initiate the process by transforming our finalMerge df to longer in order to add the classification of type to be able to group the data by “type” into a df called mergeLong. We then transform this df into a wider format to assign each year into its own column, into a df called mergeWide. Using this new df, I create separate df for obesity and population data in order to more easily manipulate the data to create our additional columns. I created a new df called mutateWide to create a df with just Area, years 2015 and 2022, and our new calculated ob_growth and pop_growth columns. To find the obesity and population growth by percentage growth, I take the value in 2022 and subtract the value in 2015 and divide by the 2015 value for each classification and each Area (state), removing NA values. Find the highest and lowest values of each obesity growth and population growth by arranging by arrange(desc()) function, and head() and tail() functions. 
For plotting, I create six total graphs to visualize the relationships between the growths and compare them. Utilizing ggplot I create a line graph with a trend line to visualize the highest/lowest obesity rate of change, as well as the highest/lowest population rate of change. To see if there is a relationship between the rates of change of obesity and population, I print the top and lowest 5 states for growth for both classifications. When I could not find the same names in each, I created a histogram to see the frequency and distribution of obesity and population growth. Using this, I can compare the other growth (obesity or population) of the highest/lowest growth states to see where their other growth lands on the distribution graph. The last two line graphs are comparing the obesity and population growths of the opposite highest and lowest states to see how they compare. 

3c)
Question 1: 
We chose to focus on recent years, 2015-2022. We figured our findings would be most beneficial to our target audience, the common public, if it was relevant. We hope our data will help those looking to find where they want to live and any other geographic decisions they might have.

Question 2: 
Our sample selection was centered on the timeframe spanning 2015 to 2022. This deliberate choice aimed to ensure our focus on pertinent and recent periods, enabling us to discuss results that remain relevant to our current context.

Question 3: 
NA

Question 4: 
Our sample selection was centered on the timeframe of years 2015-2022. We chose this sample as we wanted to see trends across the years, however, still relevant data to present. To find the growth, we should have a substantial span of time of go over. 

3d) Missing Values?
We overall did not need to remove values to attain our goals. 
3e)
NA. The only statistical methods we used were simple math and formulas. 



## Data cleaning
### Read in our data and clean population data for 2010-2019
```{r, echo = FALSE}
pop_2010_2019 <- read_delim("../Data/pop_2010_2019.csv")
short_pop_10_19 <- subset(pop_2010_2019, select = c("Area", "2015", "2016", "2017", "2018", "2019"))
```

### Read in our data and clean population data for 2020-2022
```{r, echo = FALSE}
pop_2020_2022 <- read_delim("../Data/pop_2020_2022.csv")
short_pop_20_22 <- subset(pop_2020_2022, select = c("NAME", "POPESTIMATE2020", "POPESTIMATE2021", "POPESTIMATE2022"))
colnames(short_pop_20_22) <- c("Area", "2020", "2021", "2022")
```

### Merging population data
```{r, echo = FALSE}
mergedpop <- left_join(short_pop_10_19, short_pop_20_22)
```

### Cleaning merged population data
```{r, echo = FALSE}
popLong <- pivot_longer(mergedpop,
                        !c(Area),
                        names_to="Year",
                        values_to="Population") %>%
  mutate(Year = as.numeric(sub("X", "", Year)))
```

### Read in our data and clean obesity data for 2015-2022
```{r, echo = FALSE}
obesity_data <- read_delim("../Data/National_Obesity_By_State.csv")
short_obesity_data <- subset(obesity_data, select = c("NAME", "Obesity", "Year"))
colnames(short_obesity_data) <- c("Area","Obesity", "Year")
```

### Merging population and obesity data
```{r, echo = FALSE}
finalMerge <- left_join(popLong, short_obesity_data) 
```

**4. Talk about your results**		
## Questions & Data
## 1. Are there any correlations between a state’s population size and its obesity percentage? For example, do more populous states tend to have higher or lower obesity rates?
```{r, echo = FALSE}
#Highest and lowest obesity percentage per state

ob2015_data <- finalMerge[finalMerge$Year == "2015", ]
ob2015 <- ob2015_data[, "Obesity"]

highest_obesity_state <- ob2015_data$Area[which.max(ob2015$Obesity)]
highest_obesity_percentage <- max(ob2015$Obesity)
lowest_obesity_state <- ob2015_data$Area[which.min(ob2015$Obesity)]
lowest_obesity_percentage <- min(ob2015$Obesity)

cat("The state with the highest obesity percentage in 2015 is", highest_obesity_state, "with a percentage of",  highest_obesity_percentage, "\n")
cat("The state with the lowest obesity percentage in 2015 is", lowest_obesity_state, "with a percentage of", lowest_obesity_percentage, "\n")

#trends up to 2022

highest_obesity_data <- finalMerge[finalMerge$Area == highest_obesity_state, ]
lowest_obesity_data <- finalMerge[finalMerge$Area == lowest_obesity_state, ]

calculate_annual_change <- function(state_data) {
  annual_change <- c(0, diff(state_data$Obesity) / diff(state_data$Year))
  return(annual_change)
}

highest_obesity_annual_change <- calculate_annual_change(highest_obesity_data)
lowest_obesity_annual_change <- calculate_annual_change(lowest_obesity_data)

library(ggplot2)
combined_data <- rbind(highest_obesity_data, lowest_obesity_data)
combined_annual_change <- calculate_annual_change(combined_data)
plot_data <- data.frame(
  year = rep(2015:2022, 2),
  state = rep(c(highest_obesity_state, lowest_obesity_state), each = 8),
  annual_change = c(highest_obesity_annual_change, lowest_obesity_annual_change)
)
ggplot(plot_data, aes(x = year, y = annual_change, color = state)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +  
  labs(
    title = "Annual Change in Obesity Percentage (2015-2022)",
    x = "Year",
    y = "Annual Change in Obesity Percentage",
    color = "State"
  ) +
  theme_minimal()
```
We found that Louisiana had the highest Obesity percentage in 2015 and made little to know improvements. The average rate of obesity in the state of Louisiana continued to increase by around 0.2 percent. Colorado, on the other hand, had the lowest obesity reported percentage of all 50 states, and continued to decrease their obesity percentage by around 0.67 percent on average from the years 2015 - 2022. After reviewing the graph created by Louisiana and Colorados Obesity trends, we came to the conclusion that the highest obesity percentage continued to trend upwards in obesity rates whereas the lowest continued to trend down. This information was important for us to note because it showed that patterns of obesity rates continued, no matter where on the spectrum they fell to begin with. We continued to learn about these trends in later questions.

## 2. What is the state with the highest obesity percentage, and how does its population compare to other states with lower obesity rates? Is there a relationship between high obesity and a larger or smaller population?
```{r, echo=FALSE}
high <- finalMerge %>%
  group_by(Year) %>%
  slice_max(Obesity) %>%
  ungroup()
low <- finalMerge %>%
  group_by(Year) %>%
  slice_min(Obesity) %>%
  ungroup()
```

**Highest obesity percentage per year:**

|     State | Year | Obesity Rate |
|  -------- | :------: | :----:   |
|Louisiana  |   2015   |  36.2    |
|Mississippi|   2016   |  37.3    |
|West Virginia|   2017   |  38.1    |
|West Virginia|   2018   |  39.5      |
|Mississippi|   2018   |  39.5    |
|Mississippi|   2019   |   40.8   |
|Mississippi|   2020   |  39.7    |
|West Virginia|   2021   |   40.6   |
|West Virginia|   2022   |   41.6   |

Looking at this table we can see that Mississippi, and West Virginia consistently appear among the states with the highest obesity rates. Mississippi experienced a substantial increase from 37.3% in 2016 to a peak of 40.8% in 2019, showcasing a notable upward trend. West Virginia's obesity rates also exhibited a persistent rise, reaching a peak of 41.6% in 2022.

**Lowest obesity percentage per year:**

|     State | Year | Obesity Rate |
|  -------- | :------: | :----:   |
|Colorado   |   2015   |  20.2    |
|Colorado   |   2016   |  22.3    |
|Colorado   |   2017   |  22.6    |
|Colorado   |   2018   |  23      |
|Colorado   |   2019   |  23.8    |
|District of Columbia|   2019   |   23.8   |
|Colorado   |   2020   |  24.2    |
|District of Columbia|   2021   |   24.7   |
|District of Columbia|   2022   |   21.5   |

In this table, Colorado consistently stands out as the state with the lowest obesity rates across the specified years. The District of Columbia also consistently maintains relatively low obesity rates, showcasing a stable pattern. While Colorado's rates show slight fluctuations, the District of Columbia experienced a notable decrease from 24.7% in 2021 to 21.5% in 2022. It is also interesting to note that Colorado and the District of Columbia are the only states that make up this table. 
Upon comparing these two tables, it becomes evident that the contrast between the highest and lowest obesity rates is a minimum of 20%.
Lets visualize this:


```{r, echo=FALSE}
merge_high_low <- full_join(high, low, by = join_by(Area, Year, Population, Obesity))

ggplot(merge_high_low, aes(Year, Obesity, col = Area)) +
  geom_point() +
  geom_line() +
  labs(title = "High VS Low Obesity Rates ", x="Years", y="Obesity Rates")
```

Observing the graph, it's evident that Colorado and the District of Columbia maintain relatively stable obesity rates from 2016 to 2020. In contrast, Mississippi and West Virginia consistently experience increasing rates, except for a singular peak in Louisiana in 2015. This suggests a trend where states with higher initial obesity rates are prone to ongoing increases, while states with lower initial rates tend to sustain a consistent level or even decrease. 

```{r, echo=FALSE}
ggplot(merge_high_low, aes(Year, Population, col = Area)) +
  geom_point() + 
  geom_line() +
  labs(title = "Population Growth", x="Years", y="Population")

```
Examining the graph, it's apparent that each state has sustained a relatively steady population size from 2015 to 2022. Notably, the states with the lowest obesity rates exhibit an inverse relationship in terms of population size: Colorado consistently holds the highest population, while the District of Columbia consistently has the lowest. Conversely, states with higher obesity rates tend to fall between these two extremes, with the exception of Louisiana, which maintains a relatively high population size.


## 3. Are there any significant changes in state obesity percentages over time, and how do these changes relate to changes in state populations? 

```{r, echo = FALSE}
obesity_changes <- finalMerge %>%
  group_by(Area) %>%
  summarize(obesity_change = last(Obesity) - first(Obesity))

population_changes <- finalMerge %>%
  group_by(Area) %>%
  summarize(population_change = last(Population) - first(Population))

obesity_outliers <- obesity_changes %>%
  filter(abs(scale(obesity_change)) > 2)  

population_vs_obesity_outliers <- inner_join(population_changes, obesity_outliers, by = "Area")
```
|     Area | Population Change | Obesity Change |
|  -------- | :------: | :----:   |
|Minnesota  |   235152   |  8.5    |

The data tells us that Minnesota was the only state that saw a significant increase in obesity as it relates to the increase in population. 
By answering this question, we identified Minnesota as a state with a significant increase in obesity rates, relating this change to a substantial population growth during the analyzed period (2015-2022). The population change between these years saw a big growth of 235152. Simultaneously, we could see that Minnesota also resulted in a big change in obesity rate, in which it grew 8.5 percent. This tells us that there is a perceived relationship between population growth and increased obesity rates.


## 4.Are there any states where obesity has been increasing or decreasing at a faster rate compared to changes in population size? Does the state with the highest/lowest obesity growth have the highest/lowest population growth as well and vice versa?
```{r, echo=FALSE, results = 'hide'}
mergeLong <- finalMerge %>%
  pivot_longer(!Area&!Year, names_to = "type", values_to = "data")

mergeWide <- mergeLong %>%
  pivot_wider(names_from = "Year", values_from = data)

obesity_wide <- mergeWide %>%
  filter(type == "Obesity", na.rm = TRUE)

population_wide <- mergeWide %>%
  filter(type == "Population", na.rm = TRUE)

mutateWide <- obesity_wide %>%
  filter(na.rm = TRUE) %>%
  mutate(obesity_wide, ob_growth = (((obesity_wide$"2022" - obesity_wide$'2015')/(obesity_wide$'2015'))*100), na.rm = TRUE) %>%
  mutate(population_wide, pop_growth = (((population_wide$"2022" - population_wide$'2015')/(population_wide$'2015'))*100), na.rm = TRUE) %>%
  arrange(desc(ob_growth))%>%
  select(Area, "2015", "2022", ob_growth, pop_growth)

growthDf <- mutateWide %>%
  select(Area, ob_growth, pop_growth)
colnames(growthDf)[2] <- "Obesity"
colnames(growthDf)[3] <- "Population"
growthLong <- growthDf %>%
  pivot_longer(!Area, names_to = "type", values_to = "growth_poc")

finalData <- left_join(mergeWide, growthLong)

write.csv(finalData, "finalData.csv", row.names = FALSE)

topFiveOB <- head(mutateWide, 5)
bottomSevenOB <- tail(mutateWide, 7)
```

### Obesity Growth (percentage growth)
#### Graph 1: Comparing growth of obesity of largest and smallest growth
```{r, echo=FALSE}
mergeLong %>%
  filter(type == "Obesity", Area %in% c("Minnesota", "District of Columbia"), na.rm = TRUE)%>%
  group_by(Area) %>%
  ggplot(aes(Year, data, col = Area)) +
  geom_line(size = 0.8) +
  labs(x="Years", y="Obesity from 2015-2022")+
  geom_smooth(method = "lm", se = FALSE, alpha = 0.3, size = 0.5)
```

This graph shows the visualization of the obesity percentage of the
states with the highest (Minnesota) obesity growth rate and smallest
(District of Columbia) obesity growth rate. From this graph we can see
how steep the growth of obesity percentage is for Minnesota,
comparatively, District of Columbia's obesity drops around 2021-2022.


### Population Growth (percentage growth)
```{r, echo=FALSE}
poparrange <- mutateWide %>%
    arrange(desc(pop_growth))
topFivePop <- head(poparrange, 5)
bottomFivePop <- tail(poparrange, 6)
```

#### Graph 2: Comparing growth of population of largest and smallest growth
```{r, echo=FALSE}
mergeLong %>%
  filter(type == "Population", Area %in% c("Idaho", "West Virginia"), na.rm = TRUE)%>%
  group_by(Area) %>%
  ggplot(aes(Year, data, col = Area)) +
  geom_line(size = 0.8) +
  labs(x="Years", y="Population from 2015-2022")+
  geom_smooth(method = "lm", se = FALSE, alpha = 0.8, size = 0.2)
```

This graph shows the visualization of the population of the states with
the highest (Idaho) population growth rate and smallest (West Virginia)
population growth rate. From this graph we can see that Idaho's trend is
very positive and steep, indicating a high growth, while West Virginia
shows a strong downward trend.


### Distributions (obesity & population)
#### Graph 3: Histogram: Obesity Growth Distibution
```{r, echo=FALSE}
mutateWide %>%
  filter(na.rm = TRUE)%>%
  group_by(Area) %>%
  ggplot(aes(ob_growth)) +
  geom_histogram(col = "gold1", fill = "green4") 
```

To cross reference, I will be using this graph to evaluate the
highest/lowest population growth states. For the top five, Idaho had an
obesity growth of 0.168, which is in the middle of the distribution. The
runner ups place relatively high, with Utah being in one of the highest
obesity growth. For the lower five (pop), they all are positioned
towards the lower obesity growth end of the distribution, with the
highest at 0.179, which is in the middle.


#### Graph 4: Histogram: Population Growth Distibution
```{r, echo=FALSE}
mutateWide %>%
  filter(na.rm = TRUE)%>%
  group_by(Area) %>%
  ggplot(aes(pop_growth)) +
  geom_histogram(col = "gold1", fill = "red3") 
```

To cross reference, I will be using this graph to evaluate the
highest/lowest obesity growth states and their population growth rate.
For the top five, Minnesota had a population growth of 0.0429, which is
in the middle of the distribution, also having one of the more common
population growth rates. The runner ups places are mixed with Utah and
Montana placing relatively high, and Ohio and Wisconsin placing towards
the lower middle. For the lower five (obesity), they all are positioned
towards the lower end of the population growth rate distribution.


#### _**(reference Graph 3 & 4)**_
* _Minnesota_, with the _highest obesity growth_ has a population growth of 29,394, placing it within the middle boundaries of the pop_growth histogram. 
* _District of Columbia_, with the _lowest obesity growth_ has a population growth of -450, placing it within the lower boundaries of the pop_growth histogram. 

* _Idaho_, with the _highest population growth_ has a obesity growth of 0.25, placing it within the middle lower boundaries of the ob_growth histogram. 
* _Wyoming_, with the _lowest population growth_ has a obesity growth of 0.0250, placing it within the lower boundaries of the ob_growth histogram. 

* However if we look at  the _obesity growth ranking_, the states with highest obesity growth have an _overall higher population growth_ to the states with smaller obesity growth. The same cannot be said for the population growth ranking. 


#### Graph 5: Comparing growth of obesity of largest and smallest growth with the obesity growth of the states with highest/lowest population growth. 
```{r, echo=FALSE}
mergeLong %>%
  filter(type == "Obesity", Area %in% c("Minnesota", "District of Columbia", "Idaho", "West Virginia"), na.rm = TRUE)%>%
  group_by(Area) %>%
  ggplot(aes(Year, data, col = Area)) +
  geom_line(size = 0.8) +
  labs(x="Years", y="Obesity from 2015-2022")+
  scale_color_manual(
    values = c("Minnesota" = "blue",
               "District of Columbia" = "blue4",
               "Idaho" = "orange",
               "West Virginia" = "orange3"
    )
  )+
  geom_smooth(method = "lm", se = FALSE, alpha = 0.3, size = 0.5)
```

Looking at this graph, we can see the Idaho (highest pop growth rate)
and Minnesota (highest obesity growth rate) have similar growths, with
the obesity percentage value actually being very similar as well. We
also see that West Virginia has the lowest population growth, however
has the highest Obesity percentage values overall.


#### Graph 6: Comparison of Population growth of the states with
highest/lowest population growth with the states with highest/lowest
obesity growth: 
```{r, echo=FALSE}
mergeLong %>%
  filter(type == "Population", Area %in% c("Minnesota", "District of Columbia", "Idaho", "West Virginia"), na.rm = TRUE)%>%
  group_by(Area) %>%
  ggplot(aes(Year, data, col = Area)) +
  geom_line(size = 0.8) +
  labs(x="Years", y="Population from 2015-2022")+
  scale_color_manual(
    values = c("Minnesota" = "blue",
               "District of Columbia" = "blue4",
               "Idaho" = "orange",
               "West Virginia" = "orange3"
    )
  )+
  geom_smooth(method = "lm", se = FALSE, alpha = 0.5, size = 0.2)
```

Looking at this graph, we can see that Idaho and West Virginia both have
similar population values through the years, and intersecting around
2019, however West Virginia has a negative trend. Minnesota, with the
highest obesity rate has the highest population values of the four
states, despite having a lower pop rate of change than Idaho.


#### Comparing Growth by state (percentage growth)
```{r, echo=FALSE}
growthCompare <- mutateWide %>%
  mutate(biggerOb = if_else(ob_growth > pop_growth, "True", "False"))%>%
  arrange(desc(ob_growth))
topFiveCompare <- head(growthCompare, 5)
bottomSixCompare <- tail(growthCompare, 6)

popGrowCompare <- growthCompare%>%
  arrange(desc(pop_growth))
topFiveComPop <- head(popGrowCompare, 5)
bottomSixComPop <- tail(popGrowCompare, 6)

growthCompare %>% 
  count(biggerOb == "True", na.rm = TRUE)

summaryDf <- growthCompare %>%
  select(Area, ob_growth, pop_growth, biggerOb)
summaryDf
```

#### Graph 7: Relationship Between Obesity and Population Growth Rates (%)
```{r, echo=FALSE}
ggplot(summaryDf, aes(ob_growth, pop_growth, col = Area)) +
  geom_point() +
  labs(x = "Obesity Growth Rate (%)",
        y = "Population Growth Rate (%)") +
  geom_smooth(mapping = aes(ob_growth, pop_growth, col = "green2"))
```

Looking at this graph, we can see that there is a weak positive upward
trend between obesity and population growth rate, as obesity growth rate
increases, the population growth rate is sort of likely to be higher as
well.

### Trends
..... 

If we compare the population growth of the states with highest obesity growth: 
Table 1: **Top 5 obesity growth rate:**

|     State | ob_growth | pop_growth | Obesity > Population growth | 
|  -------- | :------: | :----:       | :----:   |
|Minnesota   |   32.6%   |  4.29%    | True |
|Utah   |   29.4%   |  13.4%    | True |
|Montana   |   28.4%   |  8.97%    | True |
|Ohio   |   24.8&   |  1.19%      | True |
|Wisconsin   |   23.8%   |  2.28%    | True |

Looking at this table, we can see that Minnesota has the highest obesity
growth rate through 2015-2022, with a growth rate of 32.6%. Runners up
are Utah (29.4%), Montana (28.4%), Ohio (24.8%), and Wisconsin (23.8%).

Table 2: **Lowest 5 obesity growth rate:**

|     State | ob_growth | pop_growth |Obesity > Population growth | 
|  -------- | :------: | :----:   | :----:   |
|District of Columbia   |   -2.71%   |  -0.533%    |False |
|Alabama   |   0.562%   |  4.57%    |False |
|Illinois   |   0.649%   |  -2.15%    |True |
|Oregon   |   1.33%   |  5.59%      |False |
|Kansas   |   1.46%   |  0.967%    |True |

Looking at this table, we can see the District of Columbia has the
smallest growth rate of -2.71%, giving it the only negative growth from
2015-2022 for obesity growth. Runners up are Alabama (0.562%), Illinois
(0.649%), Oregon (1.33%), and Kansas (1.46%).

**Top 5 population growth rate:**

|     State | pop_growth | ob_growth |Obesity > Population growth | 
|  -------- | :------: | :----:   | :----:   |
|Idaho      |  17.4%    |  16.8%    |False |
|Utah    |  13.4%    |  29.4%    |True |
|Nevada    |  10.8%    |  19.5%    |True |
|Florida |  10.1%    |  20.9%    |True |
|Texas     |  9.32%    |  6.17%    |False |

Looking at this table, we can see that Idaho has the highest population
growth rate from 2015-2022, with an average rate of change (percentage
rate) 17.4%. Runners up include Utah (13.4%), Nevada (10.8%), Florida
(10.1%), and Texas (9.32%).


**Lowest 5 population growth rate:**

|     State | pop_growth | ob_growth |Obesity > Population growth | 
|  -------- | :------: | :----:   |:----:   |
|West Virginia    |  -3.63% |  16.9%    |True |
|Illinois  |  -2.15%   |  0.649%    |True |
|Mississippi    |  -1.62%    |  3.93%    |True |
|Louisiana |  -1.59%    |  3.87%    |True |
|Wyoming     |  -0.723%    |  17.9%    |True |

Here we see that West Virginia has the smallest population growth rate
with a negative value of -3.63%. Runners up include Illinois (-2.15%),
Mississippi (-1.62%), Louisiana (-1.59%), and Wyoming (-0.723%).


**Total Overview of Obesity GR > Population GR:**

|     Value | Count | 
|  -------- | :------: |
|True    |  43 |
|False  |  8   | 
|NA    |  1    | 

43 out of the 51 states (including District of Columbia) have higher obesity growth rates compared to their population growth rate, in which we can infer that states are becoming more obese compared to their population growth rate. 

### Question 4: Findings	
It is hard to draw a confident conclusion from this data, as it is hard to see a consistent strong trend, except that Utah has both the highest population growth, and the second biggest obesity growth rate. However we can see that for the five states with the highest obesity growth rate, all of their obesity growth rates are higher than their population growth rate, with the three of the bottom five with the lowest obesity growth rate having obesity growth rates that are smaller than their population growth. 
From the table with the states with the lowest population growth rate, all five have obesity growth rates higher than their population growth. For the top five population growth states, the first and fifth states have a population growth rate greater than the obesity rate, however the other three have higher obesity rates. 
From my graphs 5-6, we can see that despite having a higher growth rate, some states have higher obesity/population values than the states with highest growth. This can be taken into account. 
Referring to the tables in question 2, West Virginia ties with Mississippi for highest obesity percentage value the most across the 8 years, topping 4 times. Also, for lowest obesity percentage per year, Colorado has the highest count at 6 years, while District of Columbia tops 3 years. Despite West Virginia having the highest obesity percentage per year the most, it does not make the top 5 obesity growth rate by percentage rate, but does have the lowest population growth rate by percentage rate. However, the District of Columbia appears to have the lowest obesity rate by percentage rate across the 8 years. 
For an overall summary, 43 states have obesity growth rates greater than their population growth rate, which we can conclude that more of their population, even standardized, are becoming more obese. 	


5b) Limitations:	
Because the obesity rates among different States over the years were so close to each other, it was hard to find clear trends. Therefore, it was hard for us to make claims within our story. For example, Colorado had highest population and lowest obesity and District of Columbia had lowest population and second lowest obesity

5c) Future work direction:	
After analyzing the data, we noticed that high obesity rates in states are constantly rising and the low obesity rates are constantly remaining low. Something that might help future analysis is to find data that would discuss the habits in these states. Potentially fast food restaurants in that area, or job security percentages. Many factors can relate to obesity and would be interesting to learn more about. In looking at what other factors relate to obesity, there could be numerous that have less to do with population and more to do with geographic location. The region that high and low obesity rates are a part of could be something interesting to study, and potentially something our target audience would be interested to learn more about. Could a state in the region with the highest obesity rates be absorbed into those habits?

**6. Summary**		
Our key takeaway from analyzing the data we found is that we did not find any clear trend between high obesity with high population rates, but we did find that states with higher obesity percentage values were found among those states with lower population growth by percentage rate. 
When discovering trends among the states with the highest obesity percentage, we found that throughout the years of 2015 and 2022, Louisiana continued to rise in their obesity rates. If we could dive deeper, we would try to find the habits or actions that may contribute to this rise, however we did find that their population size was not a factor. Colorado was the lowest state in terms of obesity percentages, and continued to decrease their rates, on average, throughout the years.
When comparing the growth rate (percentage rate) of obesity and population, we found a slight upward trend where higher obesity growth rates can slightly be attributed with higher population growth rates using percentage rate. 
We also found that of the 51 states (including the District of Columbia), 43 have obesity growth rates (percentage rate) higher than population growth rates (percentage rate), indicating that increased people and populations of states are becoming obese faster than the population can grow.  

