---
title: "finalproject"
output: html_document
date: "2023-11-30"
---
```{r}
library(tidyverse)
```
# Final Project (Population vs. Obesity by State)

## Intro
### Orignial Questions: 
### 1. Are there any correlations between a state’s population size and its obesity percentage? For example, do more populous states tend to have higher or lower obesity rates?


### 2. What is the state with the highest obesity percentage, and how does its population compare to other states with lower obesity rates? Is there a relationship between high obesity and a larger or smaller population?


### 3. Are there any significant changes in state obesity percentages over time, and how do these changes relate to changes in state populations? 


### 4.Are there any states where obesity has been increasing or decreasing at a faster rate compared to changes in population size?


## Data cleaning


### Read in our data and clean population data for 2010-2019

pop_2010_2019 <- read_delim("../Data/pop_2010_2019.csv")
short_pop_10_19 <- subset(pop_2010_2019, select = c("Area", "2015", "2016", "2017", "2018", "2019"))

### Read in our data and clean population data for 2020-2022

```{r}
pop_2020_2022 <- read_delim("../Data/pop_2020_2022.csv")
short_pop_20_22 <- subset(pop_2020_2022, select = c("NAME", "POPESTIMATE2020", "POPESTIMATE2021", "POPESTIMATE2022"))
colnames(short_pop_20_22) <- c("Area", "2020", "2021", "2022")
```

### Merging population data

```{r}
mergedpop <- left_join(short_pop_10_19, short_pop_20_22)
```

### Cleaning merged population data

```{r}
popLong <- pivot_longer(mergedpop,
                        !c(Area),
                        names_to="Year",
                        values_to="Population") %>%
  mutate(Year = as.numeric(sub("X", "", Year)))
head(popLong, 4)
```

### Read in our data and clean obesity data for 2015-2022

```{r}
obesity_data <- read_delim("../Data/National_Obesity_By_State.csv")
short_obesity_data <- subset(obesity_data, select = c("NAME", "Obesity", "Year"))
colnames(short_obesity_data) <- c("Area","Obesity", "Year")
```

### Merging population and obesity data

```{r}
finalMerge <- left_join(popLong, short_obesity_data) 
```

This is our dataset, population data from 2015-2022 and obesity percentages per state: **mergedObesityPop**. 

## Questions & Data
### 1. Are there any correlations between a state’s population size and its obesity percentage? For example, do more populous states tend to have higher or lower obesity rates?
```{r}

```


### 2. What is the state with the highest obesity percentage, and how does its population compare to other states with lower obesity rates? Is there a relationship between high obesity and a larger or smaller population?
```{r}


```

Highest obesity percentage per year: 
2015: Louisiana
2016: West Virginia
2017: West Virginia
2018: West Virginia & Mississippi (tied)
2019: Mississippi
2020: Mississippi
2021: West Virginia
2022: West Virginia


### 3. Are there any significant changes in state obesity percentages over time, and how do these changes relate to changes in state populations? 
```{r}

```


### 4.Are there any states where obesity has been increasing or decreasing at a faster rate compared to changes in population size?
```{r}

```

## Graphs
```{r}
ggplot(short_obesity_data, aes(Year, Obesity, col=area)) +
  geom_line(linewidth = 0.5)+
  labs(x="Years", y="Obesity Percentage of State")
  
```
geom_smooth(method = "lm", se = FALSE, alpha = 0.5)




